---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r}
act <- read.csv("activity/activity.csv")
summary(act)
```

In fact the interval column is the time of day at 5 minute intervals using a 24
hour clock. So, combine the 2 columns into a date-time representation.
```{r}
daytime <- strptime( 
        paste( act$date, sprintf( "%04d", act$interval)), "%Y-%m-%d %H%M")
# combine new representaion of time and original data
act <- cbind( act, daytime)
```

## What is mean total number of steps taken per day?
```{r}
 per_day <- aggregate( steps ~ date, data=act, FUN=sum)
```
The mean steps taken per day is `r mean(per_day$steps)`.
The median steps taken per day is `r median(per_day$steps)`

The following plot shows the histogram of mean steps per day.
```{r}
 hist( per_day$steps)
```


## What is the average daily activity pattern?
```{r}
library(dplyr)
by_int <- group_by( act, interval)
mean_by_int <- summarize( by_int, ave = mean(steps, na.rm=TRUE))
```

The plot
```{r}
plot( mean_by_int$ave, type="l", axes=FALSE, xlab="", ylab="" )
# want time of day labels on x axis
inttimes <- factor( strftime(  act$daytime, "%H:%M") )
axis( 1, c( seq(1, 288, 48), 289),
         c( as.character(inttimes[seq(1, 288, 48)]), "24:00"))
# use default for y axis
axis(2)
# label the axes
title( xlab = "Time of Day", ylab = "Average Steps",
       main = "Daily Step Activity" )
```

## Imputing missing values
When a sample is missing, use the mean value for the associated time of day.
```{r}
# find missing observations
missing_rows = which( !complete.cases(act))
# what time of day do they correspond to
missing_tod = unique( act$interval[missing_rows])
# create new data set
newact <- act
# loop
for (k in seq_along(missing_tod) ) {
        n <- which( is.na(newact$steps) & newact$interval==missing_tod[k])
        newact$steps[n] <- rep( mean_by_int$ave[k], length(n))
        }
```

New mean, median, and histogram
## What is mean total number of steps taken per day?
```{r}
 new_per_day <- aggregate( steps ~ date, data=newact, FUN=sum)
```
The mean steps taken per day is `r mean(new_per_day$steps)`.
The median steps taken per day is `r median(new_per_day$steps)`

The following plot shows the histogram of mean steps per day.
```{r}
 #par( mfrow=c(1,2))
 #hist( per_day$steps, border="red",  ylim=c(0,40), lty=2)
 hist( new_per_day$steps, border="red", axes=FALSE, xlab="", ylab="", main="", ylim=c(0,40))
 par( new=TRUE)
 hist( per_day$steps, border="blue", ylim=c(0,40), lty=2 )
 par(new=FALSE)
```

## Are there differences in activity patterns between weekdays and weekends?
